---
title: "Effect of climate warming on the timing of autumn leaf senescence reverses at the summer solstice"
subtitle: "Experiment 2: Effects of water and nutrient availability on autumn phenology"
---

<br><br>

## 1. Load packages and functions

<details><summary>load packages</summary>

```{r, message=FALSE, warning=FALSE}
require(ggplot2)
require(tidyverse)
require(broom)
require(patchwork)
require(lubridate)
require(data.table)
require(gmodels)
require(wesanderson)
require(mgcv)
```

</details>

<details><summary>load functions</summary>

```{r, message=FALSE, warning=FALSE}
##########################################
# SPAD reading to chlorophyll conversion #
##########################################

SPADtoChl = function(SPAD){
  Chl = ifelse(SPAD==0, 0, -0.0029*(SPAD^2) + 1.175*SPAD + 3.8506)
  return(Chl)
}
#SPAD: unitless
#Chl: leaf chlorophyll in microgram / gram fresh weight
#Intercept forced to zero

#From Percival et al. 2008: The Potential of a Chlorophyll Content SPAD Meter to Quantify Nutrient Stress in Foliar Tissue of Sycamore (Acer pseudoplatanus), English Oak (Quercus robur), and European Beech (Fagus sylvatica)
```

</details>

<br>

## 2. Load data

<details>

<summary>data tables</summary>

```{r}
################
# define paths #
################

data.dir   = "/Users/consti/Desktop/PhD/Publication_material/17_Autumn_phenology_tier2/Experiment2/R_data"
output.dir = "/Users/consti/Desktop/PhD/Publication_material/17_Autumn_phenology_tier2/Experiment2/R_output"


#############
# Load data #
#############

# Moisture
moisture.data = data.frame(fread(paste(data.dir,"SoilMoisture.csv",sep="/"), header=T, sep=","))%>%
  filter(!is.na(moisture)) %>% #delete NAs
  mutate(treatment = plyr::revalue(treatment, c("A" = "High",
                                                "B" = "Interm.",
                                                "C" = "Low",
                                                "D" = "No")))

# Chlorophyll
chlorophyll.data = read.table(paste(data.dir,"Chlorophyll.csv",sep="/"), header=T, sep=",") %>%
  mutate(WaterTreatment = plyr::revalue(WaterTreatment, c("A" = "High",
                                                          "B" = "Interm.",
                                                          "C" = "Low",
                                                          "D" = "No")),
         #Convert dates to DOY
         date = dmy(date),
         #Create treatment column
         treatment = paste(WaterTreatment, NutrientTreatment, sep='_'),)
```

</details><details><summary>ggplot themes</summary>

```{r}
######################
# Define plot themes #
######################

plotTheme1 = theme(#legend.position   = "none",
  legend.background = element_rect(fill=NA, size=0.5, linetype="solid"),
  legend.text       = element_text(color="black"),
  panel.background  = element_blank(),
  axis.text         = element_text(colour = "black"),
  panel.border      = element_rect(colour = "black", fill=NA),
  axis.line         = element_line(color = "black"),
  strip.background  = element_rect(fill=NA),
  strip.text        = element_text(colour = 'black'),
  plot.title        = element_text(face="bold",hjust = 0.5))

plotTheme2 = theme(legend.position   = "none",
                   legend.background = element_rect(fill=NA, size=0.5, linetype="solid"),
                   legend.text       = element_text(color="black"),
                   panel.grid.major  = element_blank(),
                   panel.grid.minor  = element_blank(),
                   axis.text.y       = element_blank(),
                   axis.text.x       = element_text(colour = "black"),
                   axis.ticks.y      = element_blank(),
                   axis.line.y       = element_blank(),
                   panel.background  = element_blank(),
                   axis.line         = element_line(color = "black"),
                   strip.background  = element_rect(fill=NA),
                   strip.text        = element_text(colour = 'black'),
                   plot.title        = element_text(face="bold"))

plotTheme3 = theme(legend.position   = "none",
                   legend.background = element_rect(fill=NA, size=0.5, linetype="solid"),
                   legend.text       = element_text(color="black"),
                   panel.grid.major  = element_blank(),
                   panel.grid.minor  = element_blank(),
                   panel.background  = element_blank(),
                   panel.border      = element_blank(),
                   axis.text         = element_text(colour = "black"),
                   axis.line         = element_line(color = "black"),
                   strip.background  = element_rect(fill=NA),
                   strip.text        = element_text(colour = 'black'),
                   plot.title        = element_text(face="bold"))
```

</details>

<br>

## 3. Moisture data

<details><summary>Summary and plots</summary>

```{r, message=FALSE, warning=FALSE}
moisture.data = moisture.data %>%
  mutate(
    #Convert dates to DOY
    date = dmy(date),
    #order treatments
    treatment = factor(treatment, levels=c('High','Interm.','Low','No'), ordered=T)
  )


#summarize table by treatments
data.frame(moisture.data %>% 
             group_by(treatment) %>% 
              summarise(median = median(moisture),
                        mean   = ci(moisture)[1], 
                        min    = min(moisture),
                        max    = max(moisture),
                        lowCI  = ci(moisture)[2],
                        hiCI   = ci(moisture)[3], 
                        sd     = sd(moisture) ) )


#Plots
PanelA = 
  ggplot(moisture.data, aes(x=treatment, y=moisture, fill=treatment)) + 
  geom_boxplot(coef=1e30) +
  coord_cartesian(ylim = c(1.8, 39))+
  labs(x = "Treatment", y = "Soil moisture (%)") +
  scale_fill_manual(values = c('#3B9AB2','#78B7C5','#E1AF00','#F21A00'))+
  plotTheme3 +
  theme(axis.text.x = element_text(angle = 45, hjust=1))

pd  = position_dodge(0.2) 
PanelB = ggplot(moisture.data, aes(x=date, y=moisture, colour=treatment, group=treatment)) + 
  stat_summary(fun.data = "mean_se", geom="line", size=1, position=pd) +
  stat_summary(fun.data = "mean_se", size = .3, position=pd) +
  coord_cartesian(ylim = c(1.8, 39))+
  labs(x = "Date", y = "") +
  scale_color_manual(values = c('#3B9AB2','#78B7C5','#E1AF00','#F21A00'))+
  plotTheme2
```

</details>

<br>

## 4. Chlorophyll data

<details><summary>Sample sizes</summary>

```{r}
table(chlorophyll.data$WaterTreatment, chlorophyll.data$date)
```

</details>

<details><summary>Data preparation</summary>

```{r, warning=F, message=F}
#####################################################################
## Transform SPAD to chlorophyll and get relative chlorophyll content
#####################################################################


chlorophyll.data = chlorophyll.data %>%
  group_by(ID)%>%
  mutate(
    #convert SPAD readings to Chlorophyll content (Percival et al. 2008)
    chlorophyll.leaf  = SPADtoChl(SPAD),
    #add scaling factor (leaf abscission)
    chlorophyll.plant = chlorophyll.leaf*scalingFactor,
    #get relative chlorophyll content per individual
    chlorophyll.plant.rel = chlorophyll.plant / max(chlorophyll.plant), 
    #get date of maximum chlorophyll content per individual
    date.max.chl    = date[which.max(chlorophyll.plant)]) %>%
  ungroup()



##################################################################################
## Interpolate 10% and 50% senescence dates from seasonal chlorophyll measurements
##################################################################################


# Interpolate EOS10 (date chlorophyll last dropped below 90%)
senescence.data.10 = chlorophyll.data %>%
  group_by(ID, treatment,NutrientTreatment,WaterTreatment) %>%
  # Find the last row with >90% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.plant.rel > 0.9))+1,
         row_number() >= max(which(chlorophyll.plant.rel > 0.9))) %>%
  #Linear interpolation
  summarize(EOS     = as.Date(approx(chlorophyll.plant.rel, date, .9,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS10") %>%
  ungroup()

# Interpolate EOS50 (date chlorophyll last dropped below 50%)
senescence.data.50 = chlorophyll.data %>%
  group_by(ID, treatment,NutrientTreatment,WaterTreatment) %>%
  # Find the last row with >50% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.plant.rel > 0.5))+1,
         row_number() >= max(which(chlorophyll.plant.rel > 0.5))) %>%
  #Linear interpolation
  summarize(EOS     = as.Date(approx(chlorophyll.plant.rel, date, .5,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS50") %>%
  ungroup()

# Rbind
senescence.data = rbind(senescence.data.10,senescence.data.50)
```

</details>

<details><summary>Senescence date analysis</summary>

```{r,warning=F}
# Linear model
resultsLM = senescence.data %>% 
    group_by(phenophase) %>% 
    do({model = lm(EOS.DOY ~ NutrientTreatment+WaterTreatment, data=.)  # create your model
    data.frame(tidy(model),
               lowCI=ci(model)[,2],
               hiCI=ci(model)[,3])
    }) %>%
  filter(!term %in% c("(Intercept)")) %>%
  mutate(term = factor(term, levels=c("WaterTreatmentInterm.",
                                      'WaterTreatmentLow',
                                      "WaterTreatmentNo",
                                      "NutrientTreatmentLow")),
         phenophase = factor(phenophase, levels=c("EOS10","EOS50"))
                        )

as.data.frame(resultsLM) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=3)
```

</details>

<br>

## 5. Figures

</details>

<details><summary>plot 1: Senescence date boxplots</summary>

```{r, fig.align = "center"}
#order treatments
senescence.data = senescence.data %>%
  mutate(WaterTreatment    = factor(WaterTreatment,    levels=c('High','Interm.','Low','No'), ordered=T),
         NutrientTreatment = factor(NutrientTreatment, levels=c('High','Low'), ordered=T))

#Plot EOS50
SenescenceBoxplot = senescence.data %>%
  filter(phenophase == "EOS50") %>%
  ggplot(mapping = aes(x = WaterTreatment, y = EOS, fill=NutrientTreatment)) +
  geom_boxplot(position=position_dodge(0.8)) +
  xlab("Treatment") +
  ylab("EOS50") +
  scale_fill_manual(values = c("#F21A00","#E1AF00")) +
  coord_cartesian(ylim=c(as.Date(c('2020-09-01','2020-11-05'))))+
  plotTheme3

#Save PDF
pdf(paste(output.dir,"SenescenceBoxplot.pdf",sep="/"), width=3, height=3, useDingbats=FALSE)
SenescenceBoxplot
dev.off()

SenescenceBoxplot

#Plot EOS50 and EOS10
SenescenceBoxplot = senescence.data %>%
  ggplot(mapping = aes(x = WaterTreatment, y = EOS, fill=NutrientTreatment)) +
  geom_boxplot(position=position_dodge(0.8)) +
  xlab("Treatment") +
  ylab("Senescence date") +
  scale_fill_manual(values = c("#F21A00","#E1AF00")) +
  facet_wrap(~phenophase, ncol=2, scales="free_y")+
  plotTheme1 

SenescenceBoxplot
```

</details>

<details><summary>plot 2: Linear model effects</summary>

```{r, fig.align = "center"}
LMplot = ggplot(resultsLM, aes(x=term, y=estimate, fill=term)) + 
  geom_hline(yintercept = 0)+
  geom_bar(position=position_dodge(), stat="identity") +
  geom_errorbar(aes(ymin=lowCI, ymax=hiCI),
                width=.2,                    # Width of the error bars
                position=position_dodge(.9))+
  coord_cartesian(ylim=c(-22,22))+
  scale_fill_manual(values = c('#3B9AB2','#3B9AB2','#3B9AB2',"#F21A00")) +
  facet_wrap(~phenophase, ncol=2)+
  plotTheme1+
  theme(axis.text.x = element_text(angle = 45, hjust=1),
        legend.position = "none")

#Save PDF
pdf(paste(output.dir,"LinearModelPlot.pdf",sep="/"), width=6, height=6, useDingbats=FALSE)
LMplot
dev.off()

LMplot
```
</details>

<details><summary>plot 3: Chlorophyll curves (water treatment)</summary>

```{r, warning=FALSE, fig.height = 8, fig.align = "center"}
#order treatments
chlorophyll.data = chlorophyll.data %>%
  mutate(WaterTreatment = factor(WaterTreatment, levels=c('High','Interm.','Low','No'), ordered=T))

#move points .05 to the left and right
pd = position_dodge(2) 

#plot
Chlorophyll_curves_water = ggplot(chlorophyll.data, aes(x=date, y=chlorophyll.plant.rel*100, 
                             colour=WaterTreatment, group=WaterTreatment)) +
  
  geom_hline(yintercept = 50, colour="lightgrey")+
  geom_hline(yintercept = 90, colour="lightgrey")+
  geom_hline(yintercept = 10, colour="lightgrey")+
  
  geom_vline(xintercept = as.Date('2020-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-08-01'), colour="lightgrey")+
  
  stat_summary(fun.data = "mean_cl_normal", geom="line", size = 1.2, position=pd, alpha=0.7) +
  stat_summary(fun.data = "mean_cl_normal", geom="errorbar", size = 0.8, width=0, position=pd) +
  stat_summary(fun.data = "mean_cl_normal", geom="point", size = 1.2, position=pd) +
  
  scale_color_manual(values = c('#3B9AB2','#78B7C5','#E1AF00','#F21A00'))+

  coord_cartesian(ylim=c(4,96),xlim=c(as.Date(c('2020-07-25','2020-11-05'))))+
  
  xlab("Date") +
  ylab("Relative leaf chlorophyll content (%)") +
  
  facet_wrap(~NutrientTreatment, ncol=1)+
  plotTheme1 +
  theme(legend.position = "none")

#define plot layout
layout <- "
ABB
CCC
CCC"

#Merge plots
MoistPlot = PanelA + PanelB + Chlorophyll_curves_water + plot_layout(design = layout) + plot_annotation(tag_levels = 'a')&
  theme(plot.tag = element_text(face = 'bold'))

#Save PDF
pdf(paste(output.dir,"MoisturePlot.pdf",sep="/"), width=5, height=7, useDingbats=FALSE)
MoistPlot
dev.off()

MoistPlot
```

</details>

<details><summary>plot 4: Chlorophyll curves (nutrient treatment)</summary>

```{r, fig.align = "center"}
Chlorophyll_nutrients = ggplot(chlorophyll.data, aes(x=date, y=chlorophyll.plant.rel*100, 
                             colour=NutrientTreatment, group=NutrientTreatment)) +
  
  geom_hline(yintercept = 50, colour="lightgrey")+
  geom_hline(yintercept = 90, colour="lightgrey")+
  geom_hline(yintercept = 10, colour="lightgrey")+
  
  geom_vline(xintercept = as.Date('2020-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-08-01'), colour="lightgrey")+
  
  stat_summary(fun.data = "mean_cl_normal", geom="line", size = 1.2, position=pd, alpha=0.7) +
  stat_summary(fun.data = "mean_cl_normal", geom="errorbar", size = 0.8, width=0, position=pd) +
  stat_summary(fun.data = "mean_cl_normal", geom="point", size = 1.2, position=pd) +
  
  scale_colour_manual(values = c("#F21A00","#E1AF00")) +
  
  coord_cartesian(ylim=c(4,96),xlim=c(as.Date(c('2020-07-25','2020-11-05'))))+
  
  xlab("Date") +
  ylab("Relative leaf chlorophyll content (%)") +
  
  facet_wrap(~WaterTreatment, ncol=1)+
  plotTheme1

#Save PDF
pdf(paste(output.dir,"NutrientPlot.pdf",sep="/"), width=5.5, height=6, useDingbats=FALSE)
Chlorophyll_nutrients
dev.off()

Chlorophyll_nutrients
```

</details>

<details><summary>plot 5: Chlorophyll curves (Individuals)</summary>

```{r, fig.width = 10, fig.asp = 1.62, fig.align = "center"}
################################
# Relative chlorophyll content #
################################

Chlorophyll_individuals = chlorophyll.data %>%
  ggplot(aes(x=date, y=chlorophyll.plant.rel*100, 
                             colour=treatment, fill=treatment)) +
  
  geom_vline(xintercept = as.Date('2020-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-08-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-07-01'), colour="lightgrey")+
  
  geom_hline(yintercept = 50, colour="black")+
  geom_hline(yintercept = 90, colour="lightgrey")+
  geom_hline(yintercept = 10, colour="lightgrey")+
  
  geom_line(aes(x=date, y=chlorophyll.plant.rel*100), color="black")+
  geom_point()+
  geom_area(aes(x=date, y=chlorophyll.plant.rel*100), alpha=0.5) +

  geom_vline(data=senescence.data[senescence.data$phenophase=="EOS10",], 
             aes(xintercept = EOS), linetype = "dashed")+
  geom_vline(data=senescence.data[senescence.data$phenophase=="EOS50",], 
             aes(xintercept = EOS), linetype = "dashed", color="blue")+

  scale_color_manual(values = wes_palette(8, name = "Zissou1", type = "continuous"))+
  scale_fill_manual(values = wes_palette(8, name = "Zissou1", type = "continuous"))+
  
  coord_cartesian(ylim=c(4,96),xlim=c(as.Date(c('2020-07-01','2020-11-25'))))+
  
  xlab("Date") +
  ylab("Relative leaf chlorophyll content (%)") +
  
  facet_wrap(~ID, dir="v", nrow=15)+
  plotTheme1

#Save PDF
pdf(paste(output.dir,"IndividualPlot_ChlRel.pdf",sep="/"), width=9, height=18, useDingbats=FALSE)
Chlorophyll_individuals
dev.off()

Chlorophyll_individuals


################################
# Absolute chlorophyll content #
################################

Chlorophyll_absolute_individuals = chlorophyll.data %>%
  ggplot(aes(x=date, y=chlorophyll.plant, 
                             colour=treatment, fill=treatment)) +
  
  geom_vline(xintercept = as.Date('2020-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-08-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2020-07-01'), colour="lightgrey")+
  
  geom_hline(yintercept = 50, colour="lightgrey")+
  geom_hline(yintercept = 25, colour="lightgrey")+
  
  geom_line(aes(x=date, y=chlorophyll.plant), color="black")+
  geom_point()+
  geom_area(aes(x=date, y=chlorophyll.plant), alpha=0.5) +

  geom_vline(data=senescence.data[senescence.data$phenophase=="EOS10",], 
             aes(xintercept = EOS), linetype = "dashed")+
  geom_vline(data=senescence.data[senescence.data$phenophase=="EOS50",], 
             aes(xintercept = EOS), linetype = "dashed", color="blue")+

  scale_color_manual(values = wes_palette(8, name = "Zissou1", type = "continuous"))+
  scale_fill_manual(values = wes_palette(8, name = "Zissou1", type = "continuous"))+
  
  coord_cartesian(ylim=c(2,50), xlim=c(as.Date(c('2020-07-01','2020-11-05'))))+
  
  xlab("Date") +
  ylab("Leaf chlorophyll content (microgram per gram fresh weight)") +
  
  facet_wrap(~ID, dir="v", nrow=15)+
  plotTheme1

#Save PDF
pdf(paste(output.dir,"IndividualPlot_ChlAbs.pdf",sep="/"), width=8, height=18, useDingbats=FALSE)
Chlorophyll_absolute_individuals
dev.off()

Chlorophyll_absolute_individuals
```

</details>

<br>

## 6. Reproducibility

</details>

<details><summary>Reproducibility info</summary>

```{r}
## date time
Sys.time()

## session info
sessionInfo()
```

</details>